Tropical Mathematics and the Lambda Calculus

### 2023-01-22 section4.tex
commit

As morphisms in $\LREL$ correspond to tropical linear functions, morphisms in the coKleisli category $\LREL_{!}$ 
correspond to a generalization of tropical Laurent series, i.e.~the tropicalization of power series. 
The study of power series in tropical mathematics is very recent (see e.g.~\cite{Porzio2021}) and many properties of these functions still have to be well-understood.

In this section we recall some known facts about tropical Laurent series, and we establish some new properties, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.

@@

A tropical polynomial iss a piece-wise linear function of the form \begin{equation}
f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
\end{equation}
For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes. For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

 In a similar way, the tropical polynomials can be identified with the tLs
$f:\Lawv^X\to\Lawv^Y$ 
 for which the support $\C F=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ is \emph{finite}, which have thus shape $f(x)_b:=\min\limits_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
 This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.

@@

%In Example~\ref{ex:famous_ex}, 
Again, by looking at Fig~\ref{fig:plot1}, it appears that $\varphi$ behaves \emph{locally} like the polynomials $\varphi_{n}$. In particular, for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. Indeed, we can compute  by hand that $\varphi$ coincides with $\varphi_{n}$ 
for $\epsilon \geq \log2/2^{n+1}$.
However, at
%
%$x\in [\epsilon,\infty]$, with $\epsilon>0$
%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
 $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} \frac{1}{2^n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}],
%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
which actually motivated our first investigations.
In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result.

@@

\begin{theorem}\label{theorem:fepsilon}
 Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
 For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
% 
% \begin{enumerate}
%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
  %\item 
$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
% \end{enumerate}
\end{theorem}
\begin{proof}[Proof sketch]
\mbox{$\C F_\epsilon:=\set{n\in\N \mid
\widehat f(n)<\infty \textit{ and } 
\widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$}, where $\leq$ is the pointwise order on $\N^k$.
The core of the proof is showing that this set is indeed finite and enough for computing $f$.
\end{proof}

@@

The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
By considering the usual norm $\norm{x}_\infty:=\sup\limits_{a\in X} \absv{x_a}$ on $\Lawv^X$, we can generalise this property by dropping the case of $x$ having some $0$ coordinate:

@@

We conclude this subsection by noticing that $\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} (see~\cite{Selinger2004}, or the appendix, for such notions).
Its cone structure induces an order on it, called its \emph{cone order}:
$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
This makes it a Scott-continuous dcpo.
Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
We will just show that:

@@

The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.

As a first remark, let us mention that an immediate consequence of the lines after Theorem~\ref{thm:modelsTaylor} is:

\begin{corollary}\label{thm:M=infLip}
 If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is the $\inf$ of \emph{Lipschitz} functions.
\end{corollary}

Another remark is that, as consequence of Theorem~\ref{theorem:fepsilon}, the tLs with finitely many variables are locally Lipschitz on all $\BB R_{>0}$.
In this subsection we refine and generalise this kind of results.
%
%We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.

@@

Let us now consider the case of bounded exponentials:
\begin{proposition}\label{prop:boundedlip}
If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
\end{proposition}
\begin{proof}[Proof sketch]
This follows from Proposition \ref{prop:troplinear} and the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $! x$ to $\C M_{\leq n}(X)$.%
%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
\end{proof}
This result is perfectly analogous to what happens in the metric models discussed in Section \ref{section2}, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.

We can now look at what happens with tLs, i.e.~when considering the full exponential $!$.
First, since a tropical polynomial can be represented as a tropical linear function from $\Lawv^{\C M_{\mathrm{deg}(\varphi)(X)}}$ to $\Lawv$, from Proposition \ref{prop:boundedlip} we deduce:
\begin{proposition}\label{prop:polylip}
Any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
\end{proposition}

The proposition above, together with Theorem \ref{theorem:fepsilon}, can be used to show that tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz over $(0,\infty)^{X}$. Yet, we can prove a more general statement, covering also the infinitary case.

@@

The results just presented translate into the following facts about the interpretation of higher-order programs:
\begin{corollary}
For any $\lambda$-term $M$:
\begin{enumerate}
\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M^!:\model\Gamma \to \model A$ is a Lipschitz map.
\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is a locally Lipschitz map.
\end{enumerate}
\end{corollary} 
\begin{proof}
As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Prop.~\ref{prop:polylip}.
Claim 2) follows directly from Theorem~\ref{thmTLSlocLip}.
\end{proof}

Finally, let us discuss the differential structure. The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}(f):(\Lawv^{X})^{2}\to \Lawv^{Y}$, linear in its first variable, and given by 
\begin{equation}
D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu*a}+x_{a}+\mu y\right\}
\end{equation}
One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
Moreover, the Taylor formula \eqref{eq:taylorcat} yields 
a Taylor formula for tLs of the form 
$f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$.
The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
\begin{proposition}
For all tLs $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
\end{proposition}

>>>>

As morphisms in $\LREL$ correspond to tropical linear functions, morphisms in the coKleisli category $\LREL_{!}$ 
correspond to a generalization of tropical Laurent series, i.e.~the tropicalization of power series.

%many properties of these functions still have to be well-understood. 
In this section we establish some tropological and metric properties of tropical Laurent series, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
The literature on power series in tropical mathematics is very recent (see e.g.~\cite{Porzio2021}) and several results we prove in this section are, to our knowledge, new.

@@

A tropical polynomial is a piece-wise linear function of the form \begin{equation}
f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
\end{equation}
For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes. For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

 In a similar way, the tropical polynomials can be identified with the tLs
$f:\Lawv^X\to\Lawv^Y$ 
 for which the support $\C F=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ is \emph{finite}, and which have thus shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
 This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.

@@

%In Example~\ref{ex:famous_ex}, 
Again, by looking at Fig~\ref{fig:plot1}, it appears that $\varphi$ behaves \emph{locally} like the polynomials $\varphi_{n}$. In particular, for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, $\varphi$ coincides with $\varphi_{n}$ 
for $\epsilon \geq 2^{-(n+1)}$, the smallest tropical root of $\varphi_{n}$.
However, at
%
%$x\in [\epsilon,\infty]$, with $\epsilon>0$
%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
 $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} \frac{1}{2^n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}],
%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
which actually motivated our first investigations.
In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result.

@@

\begin{theorem}\label{theorem:fepsilon}
 Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
 For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
% 
% \begin{enumerate}
%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
  %\item 
$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
% \end{enumerate}
\end{theorem}
\begin{proof}[Proof sketch]
Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
The core of the proof is showing that this set is indeed finite and enough for computing $f$.
\end{proof}

@@

The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
By considering the usual norm $\norm{x}_\infty:=\sup_{a\in X} \absv{x_a}$ on $\Lawv^X$, we can generalise this property by dropping the case of $x$ having some $0$ coordinate:

@@

We conclude this subsection by noticing that $\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} (see~\cite{Selinger2004}, or the appendix, for such notions).
Its cone structure induces an order on it, called its \emph{cone order}:
$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
This makes it a Scott-continuous dcpo.
Suitable categories of cones have been recently investigated as models of probabilistic computation (\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
Here we will just show that:

@@

The norm $\norm{\cdot}_\infty$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.

@@

Let us now consider the case of bounded exponentials:
\begin{proposition}\label{prop:boundedlip}
If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
\end{proposition}
\begin{proof}[Proof sketch]
This follows from Proposition \ref{prop:troplinear} and the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $! x$ to $\C M_{\leq n}(X)$.%
%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
\end{proof}
This result is perfectly analogous to what happens in the metric models discussed in Section \ref{section2}, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.

We can now look at what happens with tLs, i.e.~when considering the full exponential $!$.
First, since a tropical polynomial can be represented as a tropical linear function from $\Lawv^{\C M_{\mathrm{deg}(\varphi)(X)}}$ to $\Lawv$, from Proposition \ref{prop:boundedlip} we deduce:
\begin{corollary}\label{prop:polylip}
Any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
\end{corollary}

Using Corollary \ref{prop:polylip}, an immediate consequence of the lines after Theorem~\ref{thm:modelsTaylor} is:

\begin{corollary}\label{thm:M=infLip}
 If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions.
\end{corollary}

% 
%
%The proposition above, together with Theorem \ref{theorem:fepsilon}, can be used to show that tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz over $(0,\infty)^{X}$.

Moreover, as consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
Actually, we can prove a more general statement, also covering the infinitary case.

@@

The results just presented translate into the following facts about the interpretation of higher-order programs:
\begin{corollary}
For any $\lambda$-term $M$:
\begin{enumerate}
\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M^!:\model\Gamma \to \model A$ is a Lipschitz map.
\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is a locally Lipschitz map.
\end{enumerate}
\end{corollary} 
\begin{proof}
As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Corollary~\ref{prop:polylip}.
Claim 2) follows directly from Theorem~\ref{thmTLSlocLip}.
\end{proof}

Finally, let us discuss the differential structure. The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by 
\begin{equation}
D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu*a}+x_{a}+\mu y\right\}
\end{equation}
One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tLs of the form 
\begin{equation}
f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}
\end{equation}
The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
\begin{proposition}
For all tLs $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
\end{proposition}

### 2023-01-22 section5C.tex
commit

\subsection{Resource Analysis for Differential Privacy}

The typical situation in differential privacy is where one considers a probabilistic query $f: \mathsf{db}\to [0,1]^{X}$ over some database, and one requires that $f$ should not be \emph{too sensitive} to small changes in the output, in other to prevent potential leaks of private information about individual items in $\mathsf{db}$ (for instance, an element $x\in\mathsf{db}$ could be the list of students of some university and $f(x)$ could indicate the percentage of female students$\mathsf{db}$).

@@

A well-studied approach to ensure differential privacy for higher-order programs is to use bounded type systems like $\mathsf{Fuzz}$ \cite{Reed2010}. Indeed, such systems come with an \emph{a priori} warrant that well-typed programs are Lipschitz.

@@

Let us equip the space $[0,1]^{X}$ with the metric $d(x,y)=\sup_{a\in X}|\log x_{a}-\log y_{a}|$. Then we have the following:
\begin{proposition}
Let $f: [0,1]^{X}\to [0,1]^{Y}$ be a function expressed by a power series. If $\trop f$ is $\epsilon$-Lipschitz over some open set $U$, then $f$ is $\epsilon$-differentially private over $\psi_{t}(U)$, for small enough $t$.
\end{proposition}
\begin{proof}[Proof sketch]
%We consider the case $X=Y=\B 1$ for simplicity. 
%Express $f$ as a power series $f(x)=\sum_{n}\psi_{e}(\widehat f_{n})x^{n}$, so that $\trop f(\alpha)= \inf_{n}\left \{
%n\alpha + \widehat f_{n}
%\right\}$.
%Let us define a family of intermediate functions $\trop_{t}f(\alpha)=\phi_{t}(\stackrel{t}{\sum_{n}}\widehat{f}_{n}\times_{t}n\alpha)$. These, by construction, satisfy
%\begin{align}\label{eq:phit}
%f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))
%\end{align}
%and, using the fact that $\alpha\sumt{t}\beta\to\min\{\alpha,\beta\}$ and $\alpha\prodt{t}\beta\to \alpha+\beta$, we can deduce that $\trop f(\alpha)=\lim_{t\to 0}\trop_{t}f(\alpha)$ (for this, we use the fact that for all $m\in \BB N$,
%$\min_{i\leq m}\alpha_{i}\leq \stackrel{t}{\sum}_{i\leq m}\alpha_{i}\leq 
%(\min_{i\leq m}\alpha_{i})+t\log m$).
We consider $X=Y=\{\star\}$ for simplicity.
With the notations from \S \ref{section3}.A, using the fact that $\lim_{t\to 0}\trop_{t}f= \trop f$, as well as the commutation $f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))$ (which follows from the definition of $\trop_{t}$), one can see that the Lipschitz condition  
$|\trop f(\phi_{t}(x))-\trop f(\phi_{t}(y))|\leq \epsilon |\phi_{t}(x)-\phi_{t}(y)|$ translates into the condition 
$f(x) \leq e^{\epsilon |\log x-\log y|} f(y)$ (supposing $y\geq x$ and $f(y)\neq 0$).
\end{proof}

>>>>

\subsection{Resource analysis for differential privacy}

The typical situation in differential privacy is where one considers a probabilistic query $f: \mathsf{db}\to [0,1]^{X}$ over some database, and one requires that $f$ should not be \emph{too sensitive} to small changes in the output, in other to prevent potential leaks of private information about individual items in $\mathsf{db}$ (for instance, an element $x\in\mathsf{db}$ could be the list of students of some university and $f(x)$ could indicate the percentage of female students in $x:\mathsf{db}$).

@@

A well-studied approach to ensure differential privacy for higher-order programs is via linear type systems with graded exponentials like $\mathsf{Fuzz}$ \cite{Reed2010}: these systems provide an \emph{a priori} warrant that well-typed programs are Lipschitz.

@@

Let us equip the space $[0,1]^{X}$ with the metric $d(x,y)=\sup_{a\in X}|\log x_{a}-\log y_{a}|$. Then we have the following:
\begin{proposition}
Let $f: [0,1]^{X}\to [0,1]^{Y}$ be a function expressed by a power series. If $\trop f$ is $\epsilon$-Lipschitz over some open set $U$, then $f$ is $\epsilon$-differentially private over $\psi_{t}(U)$, for small enough $t$.
\end{proposition}
\begin{proof}[Proof sketch]
%We consider the case $X=Y=\B 1$ for simplicity. 
%Express $f$ as a power series $f(x)=\sum_{n}\psi_{e}(\widehat f_{n})x^{n}$, so that $\trop f(\alpha)= \inf_{n}\left \{
%n\alpha + \widehat f_{n}
%\right\}$.
%Let us define a family of intermediate functions $\trop_{t}f(\alpha)=\phi_{t}(\stackrel{t}{\sum_{n}}\widehat{f}_{n}\times_{t}n\alpha)$. These, by construction, satisfy
%\begin{align}\label{eq:phit}
%f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))
%\end{align}
%and, using the fact that $\alpha\sumt{t}\beta\to\min\{\alpha,\beta\}$ and $\alpha\prodt{t}\beta\to \alpha+\beta$, we can deduce that $\trop f(\alpha)=\lim_{t\to 0}\trop_{t}f(\alpha)$ (for this, we use the fact that for all $m\in \BB N$,
%$\min_{i\leq m}\alpha_{i}\leq \stackrel{t}{\sum}_{i\leq m}\alpha_{i}\leq 
%(\min_{i\leq m}\alpha_{i})+t\log m$).
We consider $X=Y=\{\star\}$ for simplicity.
Using the fact that $\lim_{t\to 0}\trop_{t}f= \trop f$, as well as the commutation $f(\phi_{t}(\alpha)) = \phi_{t}(\trop_{t}f(\alpha))$ (which follows from the definition of $\trop_{t}$), one can see that the Lipschitz condition  
$|\trop f(\phi_{t}(x))-\trop f(\phi_{t}(y))|\leq \epsilon |\phi_{t}(x)-\phi_{t}(y)|$ translates into the condition 
$f(x) \leq e^{\epsilon d(x,y)} f(y)$ (supposing $y\geq x$ and $f(y)\neq 0$).
\end{proof}
